# R functions
## Using a function
A function does some computation on an object. The use of a function consists of:

1. A function's name;
2. Parentheses;
3. 0 or more inputs.

```{r}
Sys.Date()
sum(1:3)
```

## Function arguments
Functions can have:

- 0 arguments e.g. `Sys.Date()`;
- all mandatory arguments e.g. `length()`;
- optional arguments e.g. `sum()`;
- arguments used out of position by using their names e.g. `rnorm(mean=1, n=10).

## Common functions
Area | Functions  
------------- | -------------
Metadata | `class`, `mode`,`length`, `names`, `nrow`, `ncol` 
Measures | `mean`, `median`, `mode`, `range`, `sum`, `cumsum`
Strings | `grep`, `gsub`, `substr`, `trim`, `toupper`, `paste`
Files | `list.dirs`, `list.files`, `file.exists`


## Finding functions
Find functions by:

- Using code-completion
- Using help-pages
- Using `?mean`
- Using `??mean`
- Using `apropos("mean")`

## Seeing how functions work
You can examine how many functions work by just typing their name without any parentheses.

```{r}
Sys.Date
```

## Exercises
1. Find a function that involves Normal distributions;
2. Find some functions operate on files.


