# (APPENDIX) Appendix {-}

# Answers

## R Data Types Exercises
- Convert TRUE to an integer
```{r}
as.integer(TRUE)
```

- What is the datatype of the value returned by `Sys.time()`?
```{r}
class(Sys.time())
```

- What is the datatype of the value returned by `Sys.timezone()`?
```{r}
class(Sys.timezone())
```

- Make this quote into an R string
> "Do you think this is a game?", he said. "No I think Jenga's a game", Archer responded.

```{r}
'"Do you think this is a game?", he said. "No I think Jenga\'s a game", Archer responded.'
```

## Basic Operations Exercises
- What is the result of `pi^2`?
```{r}
pi^2
```

- Is pi greater than 3?
```{r}
pi>3
```

- Construct a statement to check if 5 is both greater than 3 and less than or equal to 6
```{r}
(5 > 3) & (5 <= 6)
```

- What are the results if you check to see if a sequence of 1 to 5 is less than or equal to 3?
```{r}
1:5 <=3
```

## R Objects Exercises
-  See what's in the built-in variable `letters`
```{r}
letters
```

-  Write a check to see if "A" is present in `letters`
```{r}
"A" %in% letters
```

-  Find out which values in the sequence 1 to 10 are greater than or equal to 3 and less than 7.
```{r}
myseq<-1:10
myseq>=3 & myseq<7
```

-  Make a vector containing the numbers 1 to 50;
```{r}
fifty<-1:50
```

-  Make a vector containing two words;
```{r}
words<-c("fifty","words")
```

-  What happens when you combine these two vectors?
```{r}
c(fifty, words)
# the numbers get converted to text
```

-  Make a data.frame using the two vectors
```{r}
fiftywords<-data.frame(fifty, words)
```

-  What happened to your text vector?
```{r}
# It got recycled 25 times
```

-  Make a list containing some of the variables you've created so far.
```{r}
list(fifty, words, fiftywords)
```

-  Return the some rows from the `iris` dataset
```{r}
head(iris)
```

## Data manipulation exercises
- Select all `LETTERS` before "X";
```{r}
LETTERS[LETTERS<"X"]
```

- Select the first 5 rows from the built-in data.frame `iris`;
```{r}
iris[1:5, ]
```

- Select the first 2 columns from `iris`;
```{r}
head(iris[,1:2])
```

- Select the column Sepal.Length from `iris` by name.
```{r}
head(iris[,"Sepal.Length"])
```

- Select rows from the `iris` data.frame where the `Sepal.Length` is greater than 5.8cm;
```{r}
head(iris[iris$Sepal.Length>5.8,])
```

- Select rows from the `iris` data.frame where the Sepal.Width is below the average for that column.
```{r}
head(iris[iris$Sepal.Width<mean(iris$Sepal.Width),])
```

- Select everything from `iris` except the Species column;
```{r}
head(iris[,-5])
```

- Create a copy of the `iris` data that just contains the first 100 rows and call it `myIris`;
```{r}
myIris<-iris[1:100,]
```

- Update the species column to the value "Unknown" in `myIris`;
```{r}
myIris$Species<-"Unknown"
```

- Delete rows from `myIris` where the sepal length is greater than 5.5.
```{r}
myIris<-myIris[myIris$Sepal.Length<=5.5,]
```


## R functions Exercises
1. Use `pmin()` to find the smallest values element-wise of the three vector `1:51`, `25:75`, `30:-20`

```{r}
pmin(1:51, 25:75, 30:-20)
```

- Use `paste()` to combine the upper case letters into a single string with `", "` between each letter

```{r}
paste(LETTERS, collapse = ", ")
```

- Use `list.files()` to see what files are in your current directory. Return the fully qualified names not just the filenames.

```{r eval=FALSE}
list.files(full.names = TRUE)
```

- View the code for `ncol()` and work out how the number of columns is being determined.

```{r}
ncol

# ncol performs the `dim()` function on the
# object we pass in. This returns a vector
# of length 2. We then subset to the second
# element, which is the number of columns.
```

## R packages Exercises
- Install `datasauRus`
```{r eval = FALSE}
install.packages("datasauRus")
```

- Load the library `datasauRus`;
```{r eval = FALSE}
library(datasauRus)
```

- Browse `datasauRus`'s help pages;
```{r eval = FALSE}
help(package="datasauRus")
```

- Read the `datasauRus` vignette;
```{r eval = FALSE}
vignette("Datasaurus")
```

