# (APPENDIX) Appendix {-}

# Answers

## R Data Types Exercises
- Convert TRUE to an integer
```{r}
as.integer(TRUE)
```

- What is the datatype of the value returned by `Sys.time()`?
```{r}
class(Sys.time())
```

- What is the datatype of the value returned by `Sys.timezone()`?
```{r}
class(Sys.timezone())
```

- Make this quote into an R string
> "Do you think this is a game?", he said. "No I think Jenga's a game", Archer responded.

```{r}
'"Do you think this is a game?", he said. "No I think Jenga\'s a game", Archer responded.'
```

## Basic Operations Exercises
- What is the result of `pi^2`?
```{r}
pi^2
```

- Is pi greater than 3?
```{r}
pi>3
```

- Construct a statement to check if 5 is both greater than 3 and less than or equal to 6
```{r}
(5 > 3) & (5 <= 6)
```

- What are the results if you check to see if a sequence of 1 to 5 is less than or equal to 3?
```{r}
1:5 <=3
```


# Software
## R
## RStudio
## Suggested packages
The following are used in example content. You do not need these to work through the exercises.

- tidyverse
- caret
- datasauRus

```{r eval=FALSE}
install.packages(c("tidyverse","caret","datasauRus"))
```
