# (APPENDIX) Appendix {-}

# Answers

## R Data Types Exercises
- Convert TRUE to an integer
```{r}
as.integer(TRUE)
```

- What is the datatype of the value returned by `Sys.time()`?
```{r}
class(Sys.time())
```

- What is the datatype of the value returned by `Sys.timezone()`?
```{r}
class(Sys.timezone())
```

- Make this quote into an R string
> "Do you think this is a game?", he said. "No I think Jenga's a game", Archer responded.

```{r}
'"Do you think this is a game?", he said. "No I think Jenga\'s a game", Archer responded.'
```

## Basic Operations Exercises
- What is the result of `pi^2`?
```{r}
pi^2
```

- Is pi greater than 3?
```{r}
pi>3
```

- Construct a statement to check if 5 is both greater than 3 and less than or equal to 6
```{r}
(5 > 3) & (5 <= 6)
```

- What are the results if you check to see if a sequence of 1 to 5 is less than or equal to 3?
```{r}
1:5 <=3
```

## R Objects Exercises
-  See what's in the variable `letters`
```{r}
letters
```

-  Write a check to see if "A" is present in `letters`
```{r}
"A" %in% letters
```

-  Find out which values in the sequence 1 to 10 are greater than or equal to 3 and less than 7.
```{r}
myseq<-1:10
myseq>=3 & myseq<7
```

-  Make a vector containing the numbers 1 to 50;
```{r}
fifty<-1:50
```

-  Make a vector containing two words;
```{r}
words<-c("fifty","words")
```

-  What happens when you combine these two vectors?
```{r}
c(fifty, words)
# the numbers get converted to text
```

-  Make a data.frame using the two vectors
```{r}
fiftywords<-data.frame(fifty, words)
```

-  What happened to your text vector?
```{r}
# It got recycled 25 times
```

-  Make a list containing some of the variables you've created so far.
```{r}
list(fifty, words, fiftywords)
```

-  Return the some rows from the `iris` dataset
```{r}
head(iris)
```

# Software
## R
## RStudio
## Suggested packages
The following are used in example content. You do not need these to work through the exercises.

- tidyverse
- caret
- datasauRus

```{r eval=FALSE}
install.packages(c("tidyverse","caret","datasauRus"))
```
